First steps towards a formalization of Forcing

### 2018-06-23 LSFA/first_steps_into_forcing.tex
mi base bibliográfica
sólo dejo mi bib porque hay repetidos
macros para markup de Isabelle
sección separada de Isa/ZF
empezando texto de preservación de pares
macros edrel, eclose; abstract; isabelle-zf es Section

\documentclass[9pt]{entcs} \usepackage{entcsmacro}
\usepackage{graphicx}
\sloppy

\input{header-draft}
\def\lastname{Gunther, Pagano, S\'anchez Terraf}
\begin{document}
\begin{frontmatter}
  \title{First steps towards a formalization of Forcing}
  \author{Emmanuel Gunther%\thanksref{ALL}
    \thanksref{myemail}}
  \address{FaMAF\\ Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina} \author{Miguel Pagano\thanksref{coemail}}
  \address{FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
  \author{Pedro S\'anchez Terraf\thanksref{co2email}}
  \address{CIEM-FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
 \thanks[myemail]{Email:
    \href{mailto:gunther@famaf.unc.edu.ar} {\texttt{\normalshape
        gunther@famaf.unc.edu.ar}}} \thanks[coemail]{Email:
    \href{mailto:pagano@famaf.unc.edu.ar} {\texttt{\normalshape
        pagano@famaf.unc.edu.ar}}}  \thanks[co2email]{Email:
    \href{mailto:sterraf@famaf.unc.edu.ar} {\texttt{\normalshape
        sterraf@famaf.unc.edu.ar}}} 
 \thanks[ALL]{Supported by Secyt-UNC project 33620180100465CB.} 
\begin{abstract} 
\end{abstract}
\begin{keyword}
Isabelle/ZF, forcing, preorder, Rasiowa-Sikorski lemma, names, generic extension.
\end{keyword}
\end{frontmatter}

@@

@@

@@

\bibliographystyle{entcs}
\bibliography{first_steps_into_forcing}
\end{document}

>>>>

\documentclass[9pt]{entcs} \usepackage{entcsmacro}
\usepackage{graphicx}
\usepackage{isabelle,isabellesym}
\sloppy

\input{header-draft}
\def\lastname{Gunther, Pagano, S\'anchez Terraf}
\begin{document}
\begin{frontmatter}
  \title{First steps towards a formalization of Forcing}
  \author{Emmanuel Gunther%\thanksref{ALL}
    \thanksref{myemail}}
  \address{FaMAF\\ Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina} \author{Miguel Pagano\thanksref{coemail}}
  \address{FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
  \author{Pedro S\'anchez Terraf\thanksref{co2email}}
  \address{CIEM-FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
 \thanks[myemail]{Email:
    \href{mailto:gunther@famaf.unc.edu.ar} {\texttt{\normalshape
        gunther@famaf.unc.edu.ar}}} \thanks[coemail]{Email:
    \href{mailto:pagano@famaf.unc.edu.ar} {\texttt{\normalshape
        pagano@famaf.unc.edu.ar}}}  \thanks[co2email]{Email:
    \href{mailto:sterraf@famaf.unc.edu.ar} {\texttt{\normalshape
        sterraf@famaf.unc.edu.ar}}} 
 \thanks[ALL]{Supported by Secyt-UNC project 33620180100465CB.} 
\begin{abstract} 
  We lay the ground for an Isabelle/ZF formalization of Cohen's technique of
  \emph{forcing}. We formalize the definition of forcing notions as
  preorders with top, dense subsets, and generic filters. We formalize
  a version of the principle of \emph{Dependent Choices} and using it
  we prove the Rasiowa-Sikorski lemma on existence of generic filters.

  Given a transitive set $M$, we define the generic extension $M[G]$,
  the canonical names for elements of $M$ and finally show that if $M$
  satisfy the axiom of pairing, the $M[G]$ also does.
\end{abstract}
\begin{keyword}
Isabelle/ZF, forcing, preorder, Rasiowa-Sikorski lemma, names, generic extension.
\end{keyword}
\end{frontmatter}

@@

\input{isabelle-zf}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{pair-gen-ext}

@@

\bibliographystyle{entcs}
\bibliography{citados}
\end{document}

### 2018-06-23 LSFA/intro.tex
macros edrel, eclose; abstract; isabelle-zf es Section

@@

The concepts of relativization and absoluteness are central to task of
transferring truth of axioms in $M$ to  $L^M$.

@@

The  Rasiowa-Sikorski lemma  states that that for every countable 
family of dense subsets of a preorder there is a filter that intersects them
all. Thus, there are generic filters $G$ for countable transitive models.  In
general, no such  $G$ belongs to $M$ and therefore the extension
$M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
in Section~\ref{sec:rasiowa-sikorski-lemma}. A preliminary result on
a version of the Axiom of Choice is formalized in
Section~\ref{sec:sequence-version-dc}.

@@

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% TeX-master: "first_steps_into_forcing"
%%% End:

>>>>

A great part of G\"odel's work on this subject has been formalized in
Isabelle by Lawrence Paulson. This paper formalizes a first part of
the machinery of forcing. In the next Subsections we discuss some of
the set-theoretical  details involved.

@@

The concepts of relativization and absoluteness are central to task of
transferring truth of axioms in $M$ to  $L^M$, and consitute the
hardest part of Paulson's development.

@@

The  Rasiowa-Sikorski lemma  states that that for every countable 
family of dense subsets of a preorder there is a filter that intersects them
all. Thus, there are generic filters $G$ for countable transitive models.  In
general, no such  $G$ belongs to $M$ and therefore the extension
$M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
in Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on
a version of the Axiom of Choice is formalized in
Section~\ref{sec:sequence-version-dc}.

@@

A central part of this formalization project involves showing that the
$\ZFC$ holds in the generic extension. This is most relevant since
forcing is essentially the only known way to \emph{extend} models of
$\ZFC$ (while preserving ordinals). The most difficult step to achieve
this goal is to define the \emph{forcing relation}, that allows to
connect satisfaction in $M$ to that of $M[G]$; this is needed to show
that the axiom schemes of Separation and Replacement hold in $M[G]$,
as well as the Powerset axiom. In this work we tackle the Pairing
Axiom. This doesn't require the 
forcing relation, but provides an illustration of the use of names.

We continue with a brief description of the Isabelle framework. 
%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% TeX-master: "first_steps_into_forcing"
%%% End:
